---
title: "Titanic"
author: "Gilbert Toroitich Tarus"
date: "14/04/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache = TRUE,warning = FALSE)
set.seed(1234)
```

## **Introduction**

This is a submission for the final project in Coursera’s Practical Machine Learning by Johns Hopkins University, third course in the Data Science: Statistics and Machine Learning Specialization.

This is a submission for the machine learning competition on [https://www.kaggle.com/](Kaggle). The aim of this competition is use machine learning to create a model that predicts which passengers survived the Titanic shipwreck.

### **Summary**

In this report, we trained five models: **Generalized Linear Model (GLM)**,**Random Forest**,**Decision Trees**, **Support Vector Machine (svm)** and **Gradient Boost Model** using k-folds cross validation for purposes of reducing noise and obtaining patterns in the training data. This models were chosen because the outcome variable is a class variable. We split the train.csv data set into training and validation sets. The test.csv data set provided was left for the purposes of the final prediction.

From the four models, the **Gradient Boosting Machines model had the highest accuracy level about 81% and a very small out of sample error about 19%**. We then use this model to do the final prediction.


### **Data**

The training data for this project are available here:


[https://www.kaggle.com/c/titanic/data?select=train.csv](https://www.kaggle.com/c/titanic/data?select=train.csv)

The test data are available here:

[https://www.kaggle.com/c/titanic/data?select=test.csv](https://www.kaggle.com/c/titanic/data?select=test.csv)

MOre information for this competition and data are found here:

[https://www.kaggle.com/c/titanic/overview](https://www.kaggle.com/c/titanic/overview)

#### **Load the required packages**

```{r packages}
library(caret);library(ggplot2);library(dplyr)
library(skimr)
library(naniar)
library(kernlab)
library(randomForest)
library(rattle)
```

## **Getting Data and cleaning data**

```{r data}
tit_train <- read.csv("data/train.csv")
tit_test <- read.csv("data/test.csv")
dim(tit_train);dim(tit_test)
```

```{r head}
view(head(tit_train[complete.cases(tit_train),],10))
```

```{r}
names(tit_train)
```
```{r glimpse}
glimpse(tit_train)
```

```{r summary}
summary(tit_train)
```


##### **Removing unnecessary and missing variables.**

For this submission, we cannot do much transformation on PassengerId, Name and Ticket variables in the data. For the purposes of modeling, we remove this variables from the data.

```{r rm}
tit_train <- tit_train %>% select(-PassengerId,-Name,-Ticket, -Cabin)
```

##### **Check for the missing values**

```{r miss_sum}
tit_train %>% miss_var_summary()
```

The variable age has 177 missing values. This is about 19.87% of the data. Since this value is not large enough to omit, we may choose to impute the missing values with the **mean** or **median**. To chose the better measure, we may have a look at the distribution of age.

```{r plot}
ggplot(tit_train,aes(x = Age))+
 geom_density()+
 geom_vline(xintercept =c(mean(tit_train$Age,na.rm = TRUE), median(tit_train$Age, na.rm = TRUE)), color= c("red","black"))
```
The two measures are close to each other. We can their choose any of them. Here we will choose to use the median to impute for the missing values.

```{r impute}
tit_train <- tit_train %>% impute_median_all()
tit_test <- tit_test %>% impute_median_all()
```


##### **Preprocessing**

```{r trsn}
# Factor variables
tit_train <- tit_train %>% 
 mutate(Pclass = as.factor(Pclass), Sex = as.factor(Sex),Parch = as.factor(Parch), Embarked = as.factor(Embarked),Survived = as.factor(Survived))
tit_test <- tit_test %>% 
 mutate(Pclass = as.factor(Pclass), Sex = as.factor(Sex),Parch = as.factor(Parch), Embarked = as.factor(Embarked))

levels(tit_train$Parch) <- c("0", "1", "2", "3", "4", "5", "6","9")
```

**Removing zero and near zero variance predictors**

```{r nzv}
nzv <- nearZeroVar(tit_train)
nzv
```
From the above output, there are no zero and near zero variance variables.

**Check for correlated data**

```{r cor}
num <- select_if(tit_train,is.numeric)
highCor<- findCorrelation(cor(num),cutoff = 0.9)
highCor
```

The ouput above shows that there are no correlated data.

## **Splitting data to training and validation sets**

We can now split the data **tit_train** to **training** and **validation** data sets. However, the test set (**"tit_test"**) will be left for the final prediction.

```{r splt}
inTrain <- createDataPartition(y=tit_train$Survived, p=0.75, list=FALSE)
training <- tit_train[inTrain,]
validation <- tit_train[-inTrain,]
```

## **Creating and Testing the Models**

We are going to fit three models: **glm**,**Random Forest**,**Decision Trees**, **gbm** and **SVM** models for classification to check which algorithm is much better to fit the data.

## **Modeling**

### **Cross validation**

To obtain the correct patterns from the data and ensure it is not getting too much noise, we use k-folds cross validation technique.

```{r trc}
train_control <- trainControl(method="cv", number=5)
```

### **Generalized Linear Model**

```{r glm}
glmFit <- train(Survived ~ ., data = training, method = "glm", family = "binomial", trControl = train_control)
# Predict probabilities
glmPred <- predict(glmFit, validation)
cmGlm <- confusionMatrix(glmPred,validation$Survived)
cmGlm
```

The GML model above as an accuracy of `r round(cmGlm$overall["Accuracy"]*100,2)`% and an out of sample error of about `r round((1-cmGlm$overall["Accuracy"])*100,2)`%. This is relative low.

### **Random Forest Model**

```{r rf}
rfMod <- train(Survived ~., data=training, method="rf", trControl = train_control, tuneLength = 5)

rfPred <- predict(rfMod, validation)
cmRF <- confusionMatrix(rfPred, validation$Survived)
cmRF
```
The random forest model above as an accuracy of `r round(cmRF$overall["Accuracy"]*100,2)`% and an out of sample error of about `r round((1-cmRF$overall["Accuracy"])*100,2)`% . This is relative high accuracy.

### **Decision Tree**

```{r Dtree}
treeMod <- train(Survived~., data=training, method="rpart", trControl = train_control, tuneLength = 5)
#Prediction
predTrees <- predict(treeMod, validation)
cmTrees <- confusionMatrix(predTrees, validation$Survived)
cmTrees
```
The random forest model above as an accuracy of `r round(cmTrees$overall["Accuracy"]*100,2)`% and an out of sample error of about `r round((1-cmTrees$overall["Accuracy"])*100,2)`% . This is relative high accuracy and it is close to that of Random Forest.


## **Support Vector Machine**

```{r svmMod}
svmMod <- train(Survived ~., data=training, method="svmRadial", trControl = train_control, tuneLength = 5, verbose = FALSE)

# Prediction
predSvm <- predict(svmMod, validation)

#Confusion matrix
cmSvm <- confusionMatrix(predSvm,validation$Survived)
cmSvm
```
The random forest model above as an accuracy of `r round(cmSvm$overall["Accuracy"]*100,2)`% and an out of sample error of about `r round((1-cmSvm$overall["Accuracy"])*100,2)`% . This is relative high accuracy and it is relatively lower than that of  Random Forest and Support Vector Machine.

## **GBM**

```{r GbmMod}
gbmMod <- train(Survived ~., data=training, method="gbm", trControl = train_control, tuneLength = 5, verbose = FALSE)

# Prediction
predGbm <- predict(gbmMod, validation)

#Confusion matrix
cmGbm <- confusionMatrix(predGbm,validation$Survived)
cmGbm
```
The random forest model above as an accuracy of `r round(cmSvm$overall["Accuracy"]*100,2)`% and an out of sample error of about `r round((1-cmSvm$overall["Accuracy"])*100,2)`% . This is relative high accuracy and it is relatively lower than that of  Random Forest and Support Vector Machine.

**Accuracy and Out of Sample Error**

```{r Acc_OSE}
GLM <- c(cmGlm$overall["Accuracy"],1-c(cmGlm$overall["Accuracy"]))
DTree <- c(cmTrees$overall["Accuracy"],1-c(cmTrees$overall["Accuracy"]))
RF <- c(cmRF$overall["Accuracy"],1-c(cmRF$overall["Accuracy"]))
SVM <- c(cmSvm$overall["Accuracy"],1-c(cmSvm$overall["Accuracy"]))
GBM <- c(cmGbm$overall["Accuracy"],1-c(cmGbm$overall["Accuracy"]))

Output <- rbind(GLM,DTree,RF,SVM,GBM)
colnames(Output) <- c("Accuracy","oo_S_Err")

Output <- Output %>% apply(.,2, round,3)
Output[order(-Output[,1]),]
```


The best model is the Gradient Boosting Machines model, with `r GBM[1]` accuracy and `r GBM[2]` out of sample error rate. We find that to be a sufficient enough model to use for our test sets.

## **Predictions on Test Set**

We will use the Gradient Boosting Machines model to do prediction on the test set since it has the highest accuracy and hence low out of sample error.

## **Gradient Boosting Machines model Predictions on the test set**

```{r PredTest}
testPredGBM <- predict(gbmMod,newdata = tit_test)
```



```{r echo=FALSE, results='hide'}
My_predictions = data.frame(PassengerId = tit_test$PassengerId,Survived = testPredGBM)
write.csv(My_predictions, "test_predictions.csv",row.names = FALSE)
```

## **Appendix**

**Feature Plot**

```{r}
featurePlot(x = tit_train[,-1], y = tit_train$Survived,plot = "pairs")
```


**correlation matrix of variables in training set**

```{r}
library(psych)
cor.plot(num,xlas = 2)
```

Plotting the models

#### **Random Forest model**

```{r}
plot(rfMod)
```

#### **Decision Trees**

```{r}
plot(treeMod)
```

```{r tree}
fancyRpartPlot(treeMod$finalModel)
```

#### **Support Vector Machine**

```{r}
plot(svmMod, plotType = "line")
```